---
title: "HW Week 5"
author: "Nolan Wark"
date: "10/12/2022"
output: github_document
---

github repo: [my github repo for homework week 5 - Nolan Wark](https://github.com/nwark/HW_Week5)

# Introduction

IF you want to know what time of day you should fly to avoid delays as much as possible depending on season, origin airport, or plane's manufacturing year.  All graphs and summary tables have "dep_delays_only" which only takes into account departure delays and "all_delays" which takes into account both departure and arrival delays.  For almost every season, origin aritport, and plane manufacturing year they fallow a similar pattern.  The pattern is almost no delays in the morning with delays growing through out the day and smaller delays after about 9-10PM to midnight.  
  
If looking at seasons winter has worse delays in the morning than any other season, and summer has the worst afternoon and night compared to the others.  If looking at origin airport, LGA has the longest time in the morning before the delays start to increase, but it also does not have a decrease in the late afternoon to midnight like the other two origins.  If looking at the maufacturing year of a plane, it seems to not have much of an effect with time of the day on delays, some years do not experience the decreasing of delays in the late afternoon to midnight and other years have too few points to try to make a claim on.

## Setup
```{r setup}
library(nycflights13)
library(dplyr)
library(ggplot2)
library(tidyr)

options(dplyr.summarise.inform = FALSE)
```

```{r}
#Data
# flights
# airlines
# airports
# planes
# unique(planes$year)
# weather
```

-

# Delays be seasons
```{r}
spring_date = as.POSIXct("2013-03-20 01:00:00", tz = "America/New_York")
summer_date = as.POSIXct("2013-06-20 01:00:00", tz = "America/New_York")
fall_date = as.POSIXct("2013-09-22 01:00:00", tz = "America/New_York")
winter_date = as.POSIXct("2013-12-21 01:00:00", tz = "America/New_York")


seasons_flights = flights %>% 
  left_join(rename(select(airports, faa, tzone), origin = faa), by="origin") %>% 
  mutate(season = ifelse(test = time_hour < spring_date, 
                         yes = "Winter", 
                         no = 
                           ifelse(test = time_hour < summer_date, 
                                  yes = "Spring",
                                  no = 
                                    ifelse(test = time_hour < fall_date,
                                           yes = "Summer",
                                           no = 
                                             ifelse(test = time_hour < winter_date, 
                                                    yes = "Fall", 
                                                    no = "Winter"))))) %>% 
  group_by(season, hour) %>% 
  summarise(dep_delays_only = mean(dep_delay, na.rm = T),
            all_delays = (sum(dep_delay, na.rm = TRUE) + sum(arr_delay, na.rm = TRUE)) / (sum(!is.na(dep_delay)) + sum(!is.na(arr_delay)))) %>% 
  arrange(factor(season, levels=c("Spring", "Summer", "Fall", "Winter")))

seasons_flights

seasons_flights %>% 
  pivot_longer(cols=c('dep_delays_only', 'all_delays'),
                names_to='delay_type',
                values_to='delay_amount') %>%
  drop_na(delay_amount) %>% 
  ggplot(aes(fill=delay_type, x=hour, y=delay_amount)) +
    geom_bar(position = 'dodge', stat='identity') +
    facet_wrap(vars(season))
```

- For all seasons, to avoid avoid delays as much as possible try to depart before about 10AM, if the season is winter try to get a plane that departs before 8AM.  For all seasons but Summer, delays quickely start to become smaller after 8PM-9PM, however for summer it stays high till the end. Fall is the best season to avoid dealays in general but all season fallow a simialar pattern of dealays becoming longer as the day progresses. 


# Delays by origin airport
```{r}
org_airport_flights = flights %>% 
  group_by(origin, hour) %>% 
  summarise(dep_delays_only = mean(dep_delay, na.rm = T),
            all_delays = (sum(dep_delay, na.rm = TRUE) + sum(arr_delay, na.rm = TRUE)) / (sum(!is.na(dep_delay)) + sum(!is.na(arr_delay))))

org_airport_flights

org_airport_flights %>% 
  pivot_longer(cols=c('dep_delays_only', 'all_delays'),
                names_to='delay_type',
                values_to='delay_amount') %>% 
  drop_na(delay_amount) %>% 
  ggplot(aes(fill=delay_type, x=hour, y=delay_amount)) +
    geom_bar(position = 'dodge', stat='identity') + 
    facet_wrap(vars(origin))
```

- All three airports fallow a similar pattern of almost no delays in the morning with the dealays becoming longer as the day progresses.  But for EWR and JFK at about 8PM the delays start to become shorter, while LGA's delays stay about the same from about 5PM to midnight.  

# Delays by plane manufactured year
```{r}
man_year_flights = flights %>% 
  left_join(rename(select(planes, tailnum, year), year_manufactured = year), by="tailnum") %>% 
  drop_na(year_manufactured) %>% 
  group_by(year_manufactured, hour) %>% 
  summarise(dep_delays_only = mean(dep_delay, na.rm = T),
            all_delays = (sum(dep_delay, na.rm = TRUE) + sum(arr_delay, na.rm = TRUE)) / (sum(!is.na(dep_delay)) + sum(!is.na(arr_delay))))

man_year_flights

man_year_flights %>% 
  pivot_longer(cols=c('dep_delays_only', 'all_delays'),
                names_to='delay_type',
                values_to='delay_amount') %>% 
  drop_na(delay_amount) %>% 
  ggplot(aes(color=delay_type, x=hour, y=delay_amount)) +
    geom_point() +
    geom_line() +
    facet_wrap(vars(year_manufactured))
  
```

- Almost all plane manuafturing years have the same trend of having delay times increase throguht the day with a small decreasing fall off around 8-9PM through midnight.  It looks like there is not as much data for planes with manufacturing years 1984 and before which is why they have jumps between points and fewer points on the chart, so it is harder to make claims about them. it looks like planes with manufacturing years of 1976, 1977, 1992, 1993, and 2010 do not have a falling of dealy times after about 9PM like most other manufacturing years.    


<!-- #Testing -->
<!-- ```{r} -->
<!-- rename(select(airports, faa, tzone), dest = faa) -->

<!-- flights$time_hour[1] < as.POSIXct("2013-01-01", tz = "America/New_York") -->
<!-- ``` -->





























